Positivity-preserving adaptive Runge-Kutta methods: results

### 2020-02-11 latex/overview.tex
Spelling mistakes and minor corrections in Paper

When computing the solution of an ODE $u ' = f(t,u) $ using a RK-Method with $s$ Stages and the Butcher tableau
\begin{align}
\begin{array}{c|c}
c &  A \\
\hline
 & b\\
\end{array}
\end{align}
the stage values $k^{(1)},\cdots,k^{(s)}$ are computed according to
\begin{equation}
k^{(j)} =  f(t^n + \Delta t c_j, u^n + \Delta t \sum_{l = 1}^{s} a_{jl} k^{(l)}  \quad j = 1,\cdots,s)
\end{equation}
Based on these values the next solution $u^{n+1}$ is computed by
\begin{equation}
u^{n+1} = u^n + \Delta t \sum_{j  = 1}^s k^{(j)} b_j .
\end{equation}
This equation can be formulated to
\begin{equation}\label{eq:Combination}
u^{n+1} = u^n + \Delta t K b
\end{equation}
where $K$ is a matrix containing the stage values $k_1,\cdots,k_s$.

The new solution $u^{n+1}$ does not comply with positifity for some cases. 
So it may be adventagious to adapt Equation\,\ref{eq:Combination} in such a way that $u^{n+1}$ is positive, instead of rejecting the Solution and computing a new set of stagevalues. This is done by adapting the weights $b$.

For the computation of the stage-values only the $A$ has to be known. The $b$ is not needed for this computation. 
Based on this Observation we can define $RKM_{A}$ as the set of RKMs that share the same $A$ and only differ by $b$.
Calculating the stage-values can be considered as solving the ODE for all RKM in $RKM_a$ at the same time.
The idea now is to choose one $RKM_{(A,b)} \in RKM_{A}$ by fixing the $b$. 
Because the stage-values are already known we can choose a $RKM_{(A,b)}$ that leads to a positive $u^{n+1}$, if $RKM_{(A,b)} \in RKM_{A} | u^{n+1} \geq 0$ exists.
When altering the $b$ the properties of an RKM are preserved. 
Especially this approach preserves linear invariants like mass conservation, because it is a regular RKM.

As it can be see in Equation\,\ref{eq:Combination} $u^{n+1}$ is a linear combination of the stage values and the previous solution $u^n$.
Because the coeffeicents $a_{jl}$ are already fixed the Order Conditions reduce to linear equations.
This means that finding the $b$ can be done by solving a linear Programming problem.

If there are no negative values in $u^{n+1}$ with the original $b$, then the method does not cause additional computational effort, because it only has to solve the LP-Problem if the regular RKM does not lead to a positive $u^{n+1}$. %(exept for the test)

@@

To illustrate the use of the method we consider the Following reaction equation from\,\cite{kopecz_comparison_2019}.

@@

The used weights $b_1,\cdots,b_5$ are also plotted in Figure\,\ref{fig:exampleI}. 
The used weights are equal for $t<1.905$. At $t=1.905$ the weights are first adapted to ensure the positivity of the solution. For $t>2.63$ the original set of weights lead to a positive solution. Because of this the used weights equal the original weights again.

@@

In the following section it is explained how the approach can be written as a LP-Problem. 
For this we have to consider which $b$ to choose. 
Generally any $b \in \mathbb{R}$ can be used. 
A first requirement would be that the new solution is positive.
We can further reduce this space by enforcing the Order Conditions. 
By incorporating the Order Conditions the space of $b$ is limited to a subspace of $\mathbb{R}$. %(Note: not linear because 0 is not insubspace)
Finally we have to give a objective function to get the optimal $b$.

@@

(Note: The primary goal here is to put the idea of changing the b into perspective. We have shown the general idea and how it works, now we want to discuss on what it does in practice. At first we state it is good to preserve the order... Get to the Example ... and show that there is more to it)

@@

We take the linear, positifity preserving ODE from \cite{kopecz_unconditionally_2018} with $u = [u_1,u_2]^T$ with the initial conditions $u_0 = [1,0]^T$ and

\begin{equation}
u'(t) = L u(t) \qquad L = \left[\begin{matrix}- a & 1\\a & -1\end{matrix}\right] \quad a =5
\end{equation}

and the first order RKM with two stages

\begin{align}
\begin{array}{c|cc}
0 &  & \\
\frac{1}{2} & \frac{1}{2} & \\
\hline
 & \frac{1}{2} & \frac{1}{2}\\
\end{array}
\end{align}

If simulated with $\Delta t = 1$ the stage values are

\begin{equation}
k^1 = \left(\begin{matrix}-5\\5\end{matrix}\right) \qquad k^2 = \left(\begin{matrix}10\\-10\end{matrix}\right)
\end{equation}

All $b$ that comply with the order Condition for the first order can be expressed as

\begin{equation}
b= \left(\begin{matrix}\frac{1}{2}\\ \frac{1}{2}\end{matrix}\right) + \alpha \left(\begin{matrix}1\\ -1\end{matrix}\right)
\end{equation}

Therefor the general solution is

\begin{equation}
u^{1} = u^0 + \Delta t  \left[k^1,k^2\right] b = \left(\begin{matrix}\frac{5}{2}\\- \frac{5}{2}\end{matrix}\right) \alpha  \left(\begin{matrix}-15\\15\end{matrix}\right)
\end{equation}

By a suitable choice of $\alpha \in [\frac{1}{6},\frac{7}{30}]$ any $u$ that complies with mass conservation and positifity can be reached. 
By adding additional constraints on the $b$ the choice of $\alpha$ can be narrowed down. 
An objective function is also needed to make the choice unique. This should be designed in a way to prefer $b$ that are close to the original weights $b_{orig}$

%This gives raise to the question weather the new solution is reasonable.
%This also poses the problem on developing a system that can decide if the new solution is has to be rejected.

@@

If the $A$ of a  preexisting RK-method is used, there is at least one solution for the linear equation system.
In order to get an optimization problem the equation system has to be underdetermined.
This implies that $\mathrm{rank}(Q) < s$. 
Because the Quadrature Conditions are linear independent, there are at least $p$ independent vectors in $Q$. This leads to the condition $p < s$.
This implies that the number of stages has to be higher as the wanted order.

@@

The second approach is to choose $b$ as a convex combination of a set of $b$s. This approach is more useful if some embedded methods with certain properties are know. 
This approach also shows more  predictable behavior because the resulting $u^{n+1}_b$ is also a convex combination of the solutions $u^{n+1}_{b_a},u^{n+1}_{b_b},\cdots$ 
This approach works best with implicit methods.
The direct adaption approach is explained in more detail in section\,\ref{sec:convex}

@@

When directly adapting the $b$ a new $b$ that complies with the Order Conditions is chosen.   
The weights $b = b_{orig} + \Delta b$ consist of the weights of the original method $b_{orig}$ and an adaption $\Delta b$. This $\Delta b \in \mathrm{ker}(Q)$. 
In Figure\,\ref{fig:b_direct} this is drawn. The red line represents the subspace of $b$ that satisfy the Order Conditions. Both $b_{orig}$ and $b$ are in this subspace.

To run it as a LP-Problem a set of constraints is constructed.  
Also a objective function is needed to determine the optimal $b$.

\subsubsection{Constraints}
The Direct Adaptation approach leads to two set of constraints on the $b$.

The first set of constraints is the positifity and can be written as 
$$u_{n+1}=u_n+\Delta K \vec{b} \geq 0$$

The second set of constraints are the Order Conditions
$$Qb=r$$

@@

\item[L1 norm] The optimisation Problem opimises for $min(|b-rkm.b|_1)$

\item[Quadrature]   The optimisation Problem opimises for $min(|b^Tq-r|)$
                            where q is the quadrature condition of the next higher order 
                            and r is the expected right hand side

\item[Hom. Order]    The optimisation Problem opimises for $min(|bTO|)$ where O are the homogenus Order Conditions of the next higher orders

@@

A reasonable property of the objective function would be that for the LP-Problem only constrained by the order condition the optimal solution is $b = b_{orig}$. 
This can be achieved by $argmin(f_{optim}(b)) = b_{orig}$. 
In practice this also means that there is no need to solve solve the optimization Problem if no constraints are violated because the solution of this problem is already known. This is adventagious for performance reasons.
The different approaches have been tested on a linear test problem from \cite{kopecz_unconditionally_2018}. The optimization problem was used at every step.
The magnitude of $b$ can get large if the Order Conditions are used as objective function.
Large values of b can lead to numerical problems. Because $\sum_{i  = 1}^s b_i = 1$, a increase of $||b||$ leads to a combination of negative and positive values. Because the computation only is done with a finite precision (and usually in floating point) large $||b||$ can lead to absorption effects.    
By changing the $b$ the stability function is altered. This can lead to oscillations if the poles of the RHS are no longer in the stability region of the resulting method.
For these reasons $||b-b_{orig}||$ seems like a good choice. The 1-Norm can be implement the with Linear Programming using slack variables. We choose

\begin{equation}
f_{obj} = ||b_{optim}-b_{orig}||_1
\end{equation}
 as objective function.

@@

If the original step $\Delta t$ was to large and there is no suitable $b$, we would like to calculate a $u^{n+1}$ using the stage values instead of rejecting the step. For this the time step is reduced after calculation the stage values. This can be done using the Order Conditions for dense Output. These can be found in \cite{hairer_runge-kutta_1993}.
We introduce $\theta \in (0,1]$. The new step taken has the length $\theta \Delta t$ whereas $\Delta t$ is the time step used when calculating the Stagevalues.
To calculate a reduced step the Order Conditions have to be adapted and a new $b_{orig}$ for the objective function (and more important the Error detection afterwards) has to be generated. 
(TODO: the $b_{orig}$ is still a open question)

@@

The dependency of $r(\theta)$ is nonlinear. This means that the maximum $\theta$ with a suitable $b$ can not be calculated by solving a LP-Problem. This means that different $\theta$ have to be tested independently. 
(Note: we could include some algorithm to search for it)

\subsection{Convex combination of existing methods}\label{sec:convex}
For some test problems (e.g. the Heat Equation solved with an implicit method of higher order), large distortions occur, even though a norm of the distortion of the b is used as a objective function. 
For implicit methods it is known that the backward euler steps lead to a positive solution for any $\Delta t$ \cite{hundsdorfer_numerical_2003}. The backward Euler is still only first Order but shows more exact results than a method of higher Order that is distorted by the adaption of the $b$, at least at some steps. 
A possible method to fix this is to generate the new b based on a convex combination of embedded methods of RKM.

\subsubsection{Constraints}
At first a set of embedded methods of the RKM $b_1,\cdots,b_d$ are needed. 
If all the used $b$ compile with the order conditions the new method is also of the same Order.
This is straightforward using

$$ Q b = Q (\chi b_{\alpha} + (1-\chi) b_{\beta}) = \chi Q  b_{\alpha} + (1-\chi) Q b_{\beta}) = \chi r + (1-\chi) r = r$$

$b_{\alpha}$ and $b_{\beta}$ are the methods that are combined to $b$ using the parameter $\chi$. The Matrix $Q$ and the vector $r$ are the Order Conditions as defined in Section\,\ref{sec:OrderCond}.

The new $b$ is defined as 
\begin{equation}\label{eq:b_convex_def}
\vec{b} = \sum_i^d a_i \vec{b_i} = Ba
\end{equation}

where $B=\left[b_1,\cdots,b_d \right]$ is a matrix containing the embedded methods.  
To ensure the that Equation\,\ref{eq:b_convex_def} yields a convex combination following constraints are enforced for the  the parameters $a,\cdots,a_d$.

\begin{equation}
 0 \leq a_1 \leq 1  \forall_{i \in \{1, \cdots d \}}
\end{equation}
\begin{equation}
 \sum_{i=1}^d a_i = 1
\end{equation}

The positifity constraint changes to

$$u_{n+1}=u_n+\Delta t K B a \geq 0$$

where $a = [a_1,\cdots,a_d]^T$

@@

As a first order method a $b$ is added that corresponds to a series of backward euler steps. These are know to be yield to a positive result.

@@

(Note: Doing dense Output here is a bit complicated, but it would not help ensure positifity only (maybe) increase exactness.  is it worth exploring this or is it better to just leave this open for further research it may be also interesting regardless the study of positifity)
Everything has downsides...

@@

In the previous sections a algorithm for choosing a $b$ has been presented. 
In the following section it is explained how this Algorithm can be integrated into the framework of existing solvers.

\subsection{Choice of Baseline Method}
An important property for the used baseline method is the Existence of embedded methods and the degrees of freedom for the b.
As noted in Section\,\ref{sec:OrderCond} the number of Stages has to be higher than the Order. 
Another important property is the number of degrees of freedom for the choice of the new weights. 
These can be calculated using $s-\mathrm{rank}(Q)$%$\mathrm{dim}(\mathrm{ker}(Q))$.
The number of degrees of freedom for different methods are shown in table\,\ref{table:DOF_exp} for explicit methods and in table\,\ref{table:DOF_imp} for implic methods.

@@

Another important property is the stability region in comparison with the maximum $\Delta t$ that ensures positifity. For this approach to be useful the RK-method has to yield negative results for $\Delta t$ smaller than the biggest stable $\Delta t$.

(Note: up to this point i do not understand why this is happening. We only see it occur depending on the Proble-RKM combination)

A third property is the existence of a positive solution. This corresponds to the question if there is a embedded first order method that ensures positivity. This is particularly interesting for Implicit methods, because the $\Delta t$ is not limited by the stability. 
If we have a embedded first Order Method we can ensure that there is always a positive solution, which might be of first Order. (Todo: Source)

@@

This gives raise to the question weather the new solution is reasonable.
This also poses the problem on developing a system that can decide if the new solution is has to be rejected.

@@

The main way is to ensure that the RKM with adapted b converges to the correct solution is to ensure that the $b$ approaches $b_{orig}$ for $\Delta t \to 0$. For this $b$ it is already know that the numeric solution converges to the exact solution.  
This is made sure by property of the Objective function that

\begin{equation}
\mathrm{argmin}(f_{optim}(b)) = b_{orig}
\end{equation}

This approximation is not very useful for real applications. 
To approximate the error of a new step we propose following approxiamtion of the local error:

@@

After adapting the $b$ the Approximation Error is checked. If $|\Delta t K(b-b_{orig})|$ is larger than the tolerance the $b$ is requected.

@@

Both errors are added to get an approximation of the total error $err = err_T +w_a err_{adapt}$. The parameter $w_a \leq 1$ is used to account for the fact that the real error is smaller than $err = err_T +err_{adapt}$. The factor also improves the stability of the step size control because $err_{adapt}$ is not as smooth as $err_T$, especially if negative values only occur for a small number of steps.
If the LP-Problem was infeasible and no $b$ was found the $err_{approx}$ is set to a custom value. This value should be chosen sufficiently large to ensure that the next step take is smaller, but still not to large to keep the stepsize control stable.

@@

%To make sure that the new adapted solution is still reasonable we measure the deviation form the Original method. In the worst case the errors add up. (todo: here some math formula... (We know that can not be the case because we already know that we will reduce the error for the quantity we will make positive, still we can use it as a upper bound.))

@@

By adapting the $b$ the used RK Method is changed. This also alters the stability function and therefore changes the region of absolute stability. 
It depends on the solved problem, the RKM and the used $\Delta t$ on how many steps the stability function is altered. For Problems where only a small number of step are affected a bigger change in the stability function would be acceptable. For Problems with require a adaptation of the weights for every step one needs to make sure that the resulting method is stable.

@@

We start with the definition of the RKM

\begin{align}
u_{n+1} &= u_n + \Delta t K (\alpha b_a+\beta b_b) \label{eq:u_n+1}  \\
u_{n+1}^{sum} &= \alpha (u_n + \Delta t K b_a) + \beta (u_n + \Delta t K b_b) \label{eq:u_n+1_sum} \\
& = (\alpha + \beta) u_n + \Delta t \alpha K b_a + \Delta t \beta K b_b \\
&= u_n + \Delta t K (\alpha b_a+\beta b_b) \\
&=u_{n+1} 
\end{align}

Now we insert the Stability functions for \ref{eq:u_n+1} and \ref{eq:u_n+1_sum} which we already know to be equal. We get

\begin{align}
u_{n+1} &= u_{n+1}^{sum} \\
R_{\alpha b_a+\beta b_b}(z) u_n &= \alpha R_{\alpha b_a}(z) u_n + \beta R_{\beta b_b}(z) u_n \\
&= (\alpha R_{\alpha b_a}(z) + \beta R_{\beta b_b}(z)) u_n
\end{align}

because this has to be true for all $u_n$ we get

@@

%(Todo:It is not straightforward if there are no pole zero cancellations in the stability function. For explicit methods the stability function is a polynomial o degree $\leq s$ (Hundsdorfer). Pole zero cancellations can not happen for these. For implicit functions that are A-stable the poles have to be in the right halve plane. Here is starts to get interesting. Because the A-Stability of one method with a certain $A$ is not a sufficient  condition that there are no Poles for all RKM with this A-Matrix. But it is possible (or easy) to prove that if $a_{ii} > 0 \forall i = 1,\cdots , s$ then there are only poles in the right halve plane where they do not matter for the further considerations.)

\subsubsection{Stability of convex combination}
If the new $b$ is chosen as a convex combination of $b$ it is easy to prove that the resulting  the Region of absolute stability is at least the Intersection of the region of absolute stability of the methods that define the convex combination. (Compare Section\,\ref{proof:convex_comb}). This also means that if all the used embedded methods used to construct the new $b$ are A-Stable, the resulting method is also A-Stable.

With this we can proof that the convex combination of two methods is absolutely stable at the points where both original methods were absolutely stable.

We want to show that $(|R_{b_a}(z)|  < 1) (|R_{b_b}(z)| < 1) \Rightarrow |R_{\chi b_a +(1- \chi) b_b}(z)| < 1$ with $\chi \in [0,1]$.
We write $\alpha = \chi$ and $\beta = 1-\chi$ for convenience.

@@

For the further discussion we are concerned with a System of $m$ coupled ODEs. 
In this setup  $u \in \mathbb{R}^m$.
Systems of ODEs with a high number of dimensions can appear when solving a PDE with the method of lines.

>>>>

When computing the solution of an ODE $u ' = f(t,u) $ using a RKM with $s$ Stages and the Butcher tableau
\begin{align}
\begin{array}{c|c}
c &  A \\
\hline
 & b\\
\end{array}
\end{align}
the stage values $y^{(1)},\cdots,y^{(s)}$ are computed according to
\begin{equation}
y^{(j)} =  u^n + \Delta t \sum_{k = 1}^{s} a_{jk} f(t^n + \Delta t c_k,y^{(k)})  \quad j = 1,\cdots,s
\end{equation}
Based on these values the next solution $u^{n+1}$ is computed by
\begin{equation}
u^{n+1} = u^n + \Delta t \sum_{j  = 1}^s f(t^n + \Delta t c_j,y^{(j)}) b_j .
\end{equation}
The stage derivatives $f^{(1)},\cdots,f^{(s)}$ can be calculated with
\begin{equation}
f^{(j)} = f(t^n + \Delta t c_j,y^{(j)})
\end{equation}
The new solution can now be calculated with
\begin{equation}\label{eq:Combination}
u^{n+1} = u^n + \Delta t F b
\end{equation}
where $F$ is a matrix containing the stage derivatives $f^{(1)},\cdots,f^{(s)}$.

The new solution $u^{n+1}$ does not comply with positivity for some cases. 
So it may be advantageous to adapt \eqref{eq:Combination} in such a way that $u^{n+1}$ is positive, instead of rejecting the Solution and computing a new set of stage values. This is done by adapting the weights $b$.

For the computation of the stage derivatives only the $A$ has to be known. The $b$ is not needed for this computation. 
Based on this Observation we can define $RKM_{A}$ as the set of RKMs that share the same $A$ and only differ in $b$.
Calculating the stage derivatives can be considered as solving the ODE for all RKM in $RKM_A$ at the same time.
The idea now is to choose one $RKM_{(A,b)} \in RKM_{A}$ by fixing the $b$. 
Because the stage derivatives are already known we can choose a $RKM_{(A,b)}$ that leads to a positive $u^{n+1}$, if $RKM_{(A,b)} \in RKM_{A} | u^{n+1} \geq 0$ exists.
When altering the $b$ the properties of an RKM are preserved. 
Especially this approach preserves linear invariants like mass conservation, because it is a regular RKM.

As it can be see in \eqref{eq:Combination} $u^{n+1}$ is a linear combination of the stage values and the previous solution $u^n$.
Because the coefficients $a_{jl}$ are already fixed the Order Conditions reduce to linear equations.
This means that finding the $b$ can be done by solving a linear Programming problem.

If there are no negative values in $u^{n+1}$ with the original $b$, then the method does not cause additional computational effort, because it only has to solve the LP-Problem if the regular RKM does not lead to a positive $u^{n+1}$. %(except for the test)

@@

To illustrate the use of the method we consider the following reaction equation from\,\cite{kopecz_comparison_2019}.

@@

The used weights $b_1,\cdots,b_5$ are also plotted in Figure\,\ref{fig:exampleI}. 
The used weights are equal the original weights for $t<1.905$. At $t=1.905$ the weights are first adapted to ensure the positivity of the solution. For $t>2.63$ the original set of weights lead to a positive solution. Because of this the used weights equal the original weights again.

@@

In the following section, it is explained how the approach can be written as a LP-Problem. 
For this we have to consider which $b$ to choose. 
Generally, any $b \in \mathbb{R}$ can be used. 
A first requirement would be that the new solution is positive.
We can further reduce this space by enforcing the Order Conditions. 
By incorporating the Order Conditions, the space of $b$ is limited to a subspace of $\mathbb{R}$. %(Note: not linear because 0 is not in subspace)
Finally, we have to give a objective function to get the optimal $b$.

@@

(Note: The primary goal here is to put the idea of changing the b into perspective. We have shown the general idea and how it works, now we want to discuss on what it does in practice. At first, we state it is good to preserve the order... Get to the Example ... and show that there is more to it)

@@

We take the linear, positivity preserving ODE from \cite{kopecz_unconditionally_2018} with $u = [u_1,u_2]^T$ with the initial conditions $u_0 = [1,0]^T$ and

\begin{equation}
u'(t) = L u(t) \qquad L = \left[\begin{matrix}- a & 1\\a & -1\end{matrix}\right] \quad a =5
\end{equation}
and the first order RKM with two stages
\begin{align}
\begin{array}{c|cc}
0 &  & \\
\frac{1}{2} & \frac{1}{2} & \\
\hline
 & \frac{1}{2} & \frac{1}{2}\\
\end{array}
\end{align}
If simulated with $\Delta t = 1$ the stage values are 
\begin{equation}
f^1 = \left(\begin{matrix}-5\\5\end{matrix}\right) \qquad f^2 = \left(\begin{matrix}10\\-10\end{matrix}\right)
\end{equation}
All $b$ that comply with the order Condition for the first order can be expressed as
\begin{equation}
b= \left(\begin{matrix}\frac{1}{2}\\ \frac{1}{2}\end{matrix}\right) + \alpha \left(\begin{matrix}1\\ -1\end{matrix}\right)
\end{equation}
With this we have one degree of freedom for the choice of $b$. This is done by the parameter $\alpha$.
If the general expression for $b$ is inserted in \eqref{eq:Combination} the general solution is 
\begin{equation}
u^{1} = u^0 + \Delta t  \left[f^1,f^2\right] b = \left(\begin{matrix}\frac{5}{2}\\- \frac{5}{2}\end{matrix}\right) \alpha  \left(\begin{matrix}-15\\15\end{matrix}\right)
\end{equation}
The solution for the original RKM would be $u^{1} = \left(\frac{5}{2} ,- \frac{5}{2}\right)^T$ which contains a negative value. But by changing the parameter $\alpha$ and therefore changing the $b$ a positive solution can be reached.
By a suitable choice of $\alpha \in \left[\frac{1}{6},\frac{7}{30}\right]$ any $u$ that complies with mass conservation and positivity can be reached. 
By adding additional constraints on the $b$ the choice of $\alpha$ can be narrowed down. 
An objective function is also needed to make the choice unique. This should be designed in a way to prefer $b$ that are close to the original weights $b_{orig}$

%This gives raise to the question whether the new solution is reasonable.
%This also poses the problem on developing a system that can decide if the new solution is has to be rejected.

@@

If the $A$ of a preexisting RK-method is used, there is at least one solution for the linear equation system.
In order to get an optimization problem, the equation system has to be underdetermined.
This implies that $\mathrm{rank}(Q) < s$. 
Because the Quadrature Conditions are linear independent, there are at least $p$ independent vectors in $Q$. This leads to the condition $p < s$.
This implies that the number of stages has to be higher as the wanted order.

@@

The second approach is to choose $b$ as a convex combination of a set of $b$s. This approach is more useful if some embedded methods with certain properties are known. 
This approach also shows more predictable behavior because the resulting $u^{n+1}_b$ is also a convex combination of the solutions $u^{n+1}_{b_a},u^{n+1}_{b_b},\cdots$ 
This approach works best with implicit methods.
The direct adaption approach is explained in more detail in section\,\ref{sec:convex}

@@

When directly adapting the $b$ a new $b$ that complies with the Order Conditions is chosen.   
The weights $b = b_{orig} + \Delta b$ consist of the weights of the original method $b_{orig}$ and an adaption $\Delta b$. The adaption is $\Delta b \in \mathrm{ker}(Q)$. 
In Figure\,\ref{fig:b_direct} this is drawn. The red line represents the subspace of $b$ that satisfy the Order Conditions. Both $b_{orig}$ and $b$ are in this subspace.

To run it as a LP-Problem a set of constraints is constructed.  
Also, an objective function is needed to determine the optimal $b$.

\subsubsection{Constraints}
The Direct Adaptation approach leads to two set of constraints on the $b$.
The first set of constraints is the positivity and can be written as 
\begin{equation}
u_{n+1}=u_n+\Delta K \vec{b} \geq 0
\end{equation}
The second set of constraints are the Order Conditions
\begin{equation}
Qb=r
\end{equation}

@@

\item[L1 norm] The optimization Problem optimizes for $min(|b-rkm.b|_1)$

\item[Quadrature]   The optimization Problem optimizes for $min(|b^Tq-r|)$
                            where q is the quadrature condition of the next higher order 
                            and r is the expected right hand side

\item[Hom. Order]    The optimization Problem optimizes for $min(|bTO|)$ where O are the homogenous Order Conditions of the next higher orders

@@

A reasonable property of the objective function would be that for the LP-Problem only constrained by the order condition the optimal solution is $b = b_{orig}$. 
This can be achieved by $argmin(f_{optim}(b)) = b_{orig}$. 
In practice, this also means that there is no need to solve solve the optimization Problem if no constraints are violated because the solution of this problem is already known. This is advantageous for performance reasons.
The different approaches have been tested on a linear test problem from \cite{kopecz_unconditionally_2018}. The optimization problem was used at every step.
The magnitude of $b$ can get large if the Order Conditions are used as objective function.
Large values of b can lead to numerical problems. Because $\sum_{i  = 1}^s b_i = 1$, an increase of $||b||$ leads to a combination of negative and positive values. Because the computation is done with a finite precision (and usually in floating point) large $||b||$ can lead to absorption effects.    
By changing the $b$ the stability function is altered. This can lead to oscillations if the poles of the RHS are no longer in the stability region of the resulting method.
For these reasons $||b-b_{orig}||$ seems like a good choice. The 1-Norm can be implement the with Linear Programming using slack variables. We choose

\begin{equation}
f_{obj} = ||b-b_{orig}||_1
\end{equation}
 as objective function.

@@

If the original step $\Delta t$ was too large and there is no suitable $b$, we would like to calculate a $u^{n+1}$ using the stage values instead of rejecting the step. For this the time step is reduced after calculation the stage values. This can be done using the Order Conditions for dense Output. These can be found in \cite{hairer_runge-kutta_1993}.
We introduce $\theta \in (0,1]$. The new step taken has the length $\theta \Delta t$ whereas $\Delta t$ is the time step used when calculating the stage values.
To calculate a reduced step the Order Conditions have to be adapted and a new $b_{orig}$ for the objective function (and more important the Error detection afterwards) has to be generated. 
(TODO: the $b_{orig}$ is still a open question)

@@

The dependency of $r(\theta)$ is nonlinear. This means that the maximum $\theta$ with a suitable $b$ cannot be calculated by solving a LP-Problem. This means that different $\theta$ have to be tested. 
(Note: we could include some algorithm to search for it)

\subsection{Convex combination of existing methods}\label{sec:convex}
For some test problems (e.g. the Heat Equation solved with an implicit method of higher order), large distortions occur, even though a norm of the distortion of the b is used as an objective function. 
For implicit methods, it is known that the backward euler steps lead to a positive solution for any $\Delta t$ \cite{hundsdorfer_numerical_2003}. The backward Euler is still only first Order but shows more exact results than a method of higher Order that is distorted by the adaption of the $b$, at least at some steps. 
A possible method to fix this is to generate the new b based on a convex combination of embedded methods of RKM.

\subsubsection{Constraints}
At first a set of embedded methods of the RKM $b_1,\cdots,b_d$ are needed. 
If all the used $b$ compile with the order conditions the new method is also of the same Order.
This is straightforward using
$$ Q b = Q (\chi b_{\alpha} + (1-\chi) b_{\beta}) = \chi Q  b_{\alpha} + (1-\chi) Q b_{\beta}) = \chi r + (1-\chi) r = r$$
$b_{\alpha}$ and $b_{\beta}$ are the methods that are combined to $b$ using the parameter $\chi$. The Matrix $Q$ and the vector $r$ are the Order Conditions as defined in Section\,\ref{sec:OrderCond}.

The new $b$ is defined as 
\begin{equation}\label{eq:b_convex_def}
b = \sum_i^d a_i b_i = Ba
\end{equation}
where $B=\left[b_1,\cdots,b_d \right]$ is a matrix containing the embedded methods.  
To ensure the that \eqref{eq:b_convex_def} yields a convex combination following constraints are enforced for the  the parameters $a,\cdots,a_d$.
\begin{equation}
 0 \leq a_1 \leq 1  \forall {i \in \{1, \cdots d \}}
\end{equation}
\begin{equation}
 \sum_{i=1}^d a_i = 1
\end{equation}
The positivity constraint changes to 
$$u_{n+1}=u_n+\Delta t K B a \geq 0$$
where $a = [a_1,\cdots,a_d]^T$

@@

As a first order method, a $b$ is added that corresponds to a series of backward euler steps. These are known to be yield to a positive result.

@@

(Note: Doing dense Output here is a bit complicated, but it would not help ensure positivity only (maybe) increase exactness.  is it worth exploring this or is it better to just leave this open for further research it may be also interesting regardless the study of positivity)
Everything has downsides...

@@

In the previous sections an algorithm for choosing a $b$ has been presented. 
In the following section, it is explained how this Algorithm can be integrated into the framework of existing solvers.

\subsection{Choice of Baseline Method}
An important property for the used baseline method is the existence of embedded methods and the degrees of freedom for the b.
As noted in Section\,\ref{sec:OrderCond} the number of Stages has to be higher than the order. 
Another important property is the number of degrees of freedom for the choice of the new weights. 
These can be calculated using $s-\mathrm{rank}(Q)$%$\mathrm{dim}(\mathrm{ker}(Q))$.
The number of degrees of freedom for different methods are shown in table\,\ref{table:DOF_exp} for explicit methods and in table\,\ref{table:DOF_imp} for implicit methods.

@@

Another important property is the stability region in comparison with the maximum $\Delta t$ that ensures positivity. For this approach to be useful the RK-method has to yield negative results for $\Delta t$ smaller than the biggest stable $\Delta t$.

(Note: up to this point I do not understand why this is happening. We only see it occur depending on the Problem-RKM combination)

A third property is the existence of a positive solution. This corresponds to the question if there is an embedded first order method that ensures positivity. This is particularly interesting for Implicit methods, because the $\Delta t$ is not limited by the stability. 
If we have an embedded first Order Method we can ensure that there is always a positive solution, which might be of first Order.

@@

This gives raise to the question whether the new solution is reasonable.
This also poses the problem on developing a system that can decide if the new solution is has to be rejected.

@@

The main way is to ensure that the RKM with adapted b converges to the correct solution is to ensure that the $b$ approaches $b_{orig}$ for $\Delta t \to 0$. For this $b$ it is already known that the numeric solution converges to the exact solution.  
This is made sure by property of the Objective function that 
\begin{equation}
\mathrm{argmin}(f_{optim}(b)) = b_{orig}
\end{equation}
This approximation is not very useful for real applications. 
To approximate the error of a new step we propose following approximation of the local error:

@@

After adapting the $b$ the Approximation Error is checked. If $|\Delta t K(b-b_{orig})|$ is larger than the tolerance the $b$ is rejected.

@@

Both errors are added to get an approximation of the total error $err = err_T +w_a err_{adapt}$. The parameter $w_a \leq 1$ is used to account for the fact that the real error is smaller than $err = err_T +err_{adapt}$. The factor also improves the stability of the step size control because $err_{adapt}$ is not as smooth as $err_T$, especially if negative values only occur for a small number of steps.
If the LP-Problem was infeasible and no $b$ was found the $err_{approx}$ is set to a custom value. This value should be chosen sufficiently large to ensure that the next step take is smaller, but still not too large to keep the step size control stable.

@@

%To make sure that the new adapted solution is still reasonable we measure the deviation from the Original method. In the worst case the errors add up. (todo: here some math formula... (We know that cannot be the case because we already know that we will reduce the error for the quantity we will make positive, still we can use it as a upper bound.))

@@

By adapting the $b$ the used RK Method is changed. This also alters the stability function and therefore changes the region of absolute stability. 
It depends on the solved problem, the RKM and the used $\Delta t$ on how many steps the stability function is altered. For Problems where only a small number of step are affected a bigger change in the stability function would be acceptable. For Problems with require an adaptation of the weights for every step one needs to make sure that the resulting method is stable.

@@

We start with the definition of the RKM
\begin{align}
u_{n+1} &= u_n + \Delta t K (\alpha b_a+\beta b_b) \label{eq:u_n+1}  \\
u_{n+1}^{sum} &= \alpha (u_n + \Delta t K b_a) + \beta (u_n + \Delta t K b_b) \label{eq:u_n+1_sum} \\
& = (\alpha + \beta) u_n + \Delta t \alpha K b_a + \Delta t \beta K b_b \\
&= u_n + \Delta t K (\alpha b_a+\beta b_b) \\
&=u_{n+1} 
\end{align}
Now we insert the Stability functions for \ref{eq:u_n+1} and \ref{eq:u_n+1_sum} which we already know to be equal. We get 
\begin{align}
u_{n+1} &= u_{n+1}^{sum} \\
R_{\alpha b_a+\beta b_b}(z) u_n &= \alpha R_{\alpha b_a}(z) u_n + \beta R_{\beta b_b}(z) u_n \\
&= (\alpha R_{\alpha b_a}(z) + \beta R_{\beta b_b}(z)) u_n
\end{align}
because this has to be true for all $u_n$ we get

@@

%(Todo: It is not straightforward if there are no pole zero cancellations in the stability function. For explicit methods, the stability function is a polynomial o degree $\leq s$ (Hundsdorfer). Pole zero cancellations cannot happen for these. For implicit functions that are A-stable the poles have to be in the right halve plane. Here is starts to get interesting. Because the A-Stability of one method with a certain $A$ is not a sufficient condition that there are no Poles for all RKM with this A-Matrix. But it is possible (or easy) to prove that if $a_{ii} > 0 \forall i = 1,\cdots , s$ then there are only poles in the right halve plane where they do not matter for the further considerations.)

\subsubsection{Stability of convex combination}
If the new $b$ is chosen as a convex combination of $b$ it is easy to prove that the resulting region of absolute stability is at least the Intersection of the regions of absolute stability of the methods that define the convex combination. This also means that if all the used embedded methods used to construct the new $b$ are A-Stable, the resulting method is also A-Stable.

We want to proof that the convex combination of two methods is absolutely stable at the points where both original methods were absolutely stable.
This can be formulated as $(|R_{b_a}(z)|  < 1) (|R_{b_b}(z)| < 1) \Rightarrow |R_{\chi b_a +(1- \chi) b_b}(z)| < 1$ with $\chi \in [0,1]$.
We write $\alpha = \chi$ and $\beta = 1-\chi$ for convenience.

@@

For the further discussion, we are concerned with a System of $m$ coupled ODEs. 
In this setup $u \in \mathbb{R}^m$.
Systems of ODEs with a high number of dimensions can appear when solving a PDE with the method of lines.
